Revealing transient strain in geodetic data with Gaussian process regression

### 2017-07-17 RevisedSubmission/hines_hetland_2017_gji.tex
added more

\subsection{Transient Strain Rates}\label{sec:Results} 
Having established a noise model and a prior for transient displacements, we use the cleaned GNSS dataset to calculate transient strain rates in the Puget Sound region.  We calculate transient strain rates for each day from January 1, 2010 to May 15, 2017. The strain rates are estimates at a grid of points spanning the study area. In Figure \ref{fig:StrainMap} we show the transient strain rates on January 1, 2016, which coincides with the height of an SSE. We have included an animation showing the map view of strain rates through time as supplementary material. The strain rates shown in Figure \ref{fig:StrainMap} are generally similar to the strain rates for the other six SSEs considered in this study. The SSEs cause compression in the Olympic Peninsula and extension east of Puget Sound. For comparison, estimated secular strain rates indicate trench perpendicular compression throughout this study region \citep{Murray2000,McCaffrey2007,McCaffrey2013}. The SSEs are thus concentrating tectonically accumulated strain energy trench-ward, and presumably pushing the subduction zone closer to failure. Similar conclusions have been drawn based on fault slip models \citep[e.g.,][]{Dragert2001,Wech2009,Schmidt2010}, which reveal that SSEs are occurring down-dip of the seismogenic zone and migrating stress upward. A key difference between the strain inferred here and strain that can be derived from fault slip models is that our estimated strain rates are not based on an assumed physical model. In contrast, fault slip models can be biased by errors in the assumed fault geometry or lithospheric rheology. Moreover, the degrees of freedom in fault slip models usually cannot be constrained by GNSS data alone, and it is necessary to impose regularization which further biases the results. Since our estimated strain rates lack such systematic errors, we can be more confident that our solution is unbiased and has meaningful uncertainties.

@@

The results we have presented thus far indicate that we are identifying the strain that we should expect to see. There are, however, features in our estimated transient strain rates which we were not expecting. For example, there is a brief period of east-west extension on the Olympic peninsula several days prior to some of the SSEs. This feature can be seen before the summer 2012 and winter 2015-2016 SSEs in Figure \ref{fig:StrainTs} as well as in the supplementary animation. A discussion on the mechanisms causing this deformation is outside the scope of this study.

\section{Discussion}\label{sec:Discussion}
We have demonstrated that transient strain rates estimated with the method described in Section \ref{sec:Method} can be used to detect SSEs and will be robust to detect other transient geophysical phenomena. Another potential application would be to use GNSS derived transient strain rates to better understand the data from borehole strain meters (BSMs). The Plate Boundary Observatory maintains 43 BSMs in Cascadia, and it has been demonstrated that BSMs are able to record transient geophysical events such as SSEs \citep[e.g.,][]{Dragert2011}. However, there are complications that prevent BSM data from being used quantitatively in geophysical studies. One difficulty is that BSM data should be calibrated with a well known strain source, such as diurnal and semidiurnal tides \citep{Hart1996,Roeloffs2010,Hodgkinson2013}. Unfortunately, the tidal forces at BSMs which record SSEs are strongly influenced by local bodies of water such as the Straight of Juan de Fuca, making it difficult to form a theoretical prediction of tidal strains \citep{Roeloffs2010}. Another complication is that noise in BSM data is not well understood. The noise consists, in part, of a long-term decay resulting from the instrument equilibrating with the surrounding rock \citep{Gladwin1987}. Typically, this noise is dealt with in an ad-hoc manner by fitting and removing exponentials and low-order polynomials. We envision that the GNSS derived strain rates from this paper can be used as a reference strain for calibrating BSM data and quantify its noise.

There is potential for a more thorough analysis of the spatio-temporal noise in GNSS data, $\eta$, than what was performed in Section \ref{sec:NoiseModel}. We did not explore the spatial covariance of $\eta$. Spatially correlated common mode noise, which results from factors such as reference frame error, is a non-negligible component of GNSS data. We ignore common mode error in this study because transients strain rates are insensitive to it. For other geophysical studies based on GNSS data, such as fault slip inversions, it may be necessary to incorporate a spatially covarying noise model \citep[e.g.,][]{Miyazaki2003}. We can also improve upon the seasonal model used in this study, which consists of four spatially uncorrelated sinusoids for each station. We did not explore the spatial covariance of seasonal deformation or the temporal roughness (i.e., the number of sinusoids needed to describe the observations). The periodic Gaussian process \citep{Mackay1998} is an alternative model for seasonal deformation and is well suited for exploring the roughness of seasonal deformation.  The periodic Gaussian process has zero mean and the covariance function
\begin{equation}\label{eq:Periodic}
T(t,t') = \phi^2 \exp\left(\frac{-\sin(\pi|t - t'|)^2}{2\tau^2}\right).
\end{equation}
Realizations have annual periodicity and the roughness is controlled by $\tau$. Decreasing $\tau$ has the same effect as including higher frequency sinusoids in the seasonal model. The optimal value for $\tau$ can be found with the REML method as described in Section \ref{sec:NoiseModel}.

>>>>

\subsection{Transient Strain Rates}\label{sec:Results} 
Having established a noise model and a prior for transient displacements, we use the cleaned GNSS dataset to calculate transient strain rates in the Puget Sound region.  We calculate transient strain rates for each day from January 1, 2010 to May 15, 2017. The strain rates are estimates at a grid of points spanning the study area. In Figure \ref{fig:StrainMap} we show the transient strain rates on January 1, 2016, which coincides with the height of an SSE. We have included an animation showing the map view of strain rates through time as supplementary material. The strain rates shown in Figure \ref{fig:StrainMap} are generally similar to the strain rates during the other six SSEs considered in this study. The SSEs cause trench perpendicular compression in the Olympic Peninsula and extension east of Puget Sound. The strain transitions from compressional to extensional strain around the southern tip of Vancouver Island, which coincides with the location of where thrust slip tends to be inferred for SSEs in the Puget Sound region \citep[e.g.,][]{Dragert2001,Wech2009,Schmidt2010}. Thus, this pattern of strain is to be expected. During the period in between SSEs, secular strain rates indicate trench perpendicular compression throughout this study region \citep{Murray2000,McCaffrey2007,McCaffrey2013}. When comparing inferred strain rates from SSEs to the secular strain rates, we see that SSEs are concentrating tectonically accumulated strain energy towards the trench, and presumably pushing the subduction zone closer to failure.

%A key difference between the strain inferred here and strain that can be derived from fault slip models is that our estimated strain rates are not based on an assumed physical model. In contrast, fault slip models can be biased by errors in the assumed fault geometry or lithospheric rheology. Moreover, the degrees of freedom in fault slip models usually cannot be constrained by GNSS data alone, and it is necessary to impose regularization which further biases the results. Since our estimated strain rates lack such systematic errors, we can be more confident that our solution is unbiased and has meaningful uncertainties.

@@

The results we have presented thus far indicate that we are identifying the strain that we should expect to see. There are, however, subtle features in our estimated transient strain rates which we were not expecting. For example, there is a brief period of east-west extension on the Olympic Peninsula several days prior to some of the SSEs. This feature can be seen before the summer 2012 and winter 2015-2016 SSEs in Figure \ref{fig:StrainTs} as well as in the supplementary animation. While this deformation is noteworthy, a discussion on the mechanisms causing it is outside the scope of this study.

\section{Discussion}\label{sec:Discussion}
Our results demonstrate that GPR is an effective tool for estimating transient strain from GNSS data, which can then be used to detect geophysical processes. Alternatively, one could detect geophysical signal by merely inspecting raw GNSS displacement time series. Indeed, the SSEs identified in Figure \ref{fig:StrainMag} do produce visually discernible displacements in the GNSS time series. However, the GNSS data also contains outliers and non-tectonic deformation that is localized to individual stations. In contrast, our estimates of transient strain only identify features that are sufficiently spatially and temporally coherent, based on our chosen prior model. Furthermore, our estimates of transient strain are insensitive to common mode noise, which is highly spatially correlated noise resulting from factors such as reference frame error. Common mode noise can obscure geophysical signal in the GNSS data, but it gets canceled out when computing the spatial gradients to obtain transient strain. Lastly, if we want to detect transient geophysical signal for the purpose of assessing seismic hazard, then strain is perhaps a more meaningful metric to monitor than displacements.

In addition to detecting geophysical processes, the GNSS derived transient strain rates can be used to better understand the data from borehole strain meters (BSMs). The Plate Boundary Observatory maintains 43 BSMs in Cascadia, and it has been demonstrated that BSMs are able to record transient geophysical events such as SSEs \citep[e.g.,][]{Dragert2011}. However, there are complications that prevent BSM data from being used quantitatively in geophysical studies. One difficulty is that BSM data should be calibrated with a well known strain source, such as diurnal and semidiurnal tides \citep{Hart1996,Roeloffs2010,Hodgkinson2013}. Unfortunately, the tidal forces at BSMs which record SSEs are strongly influenced by local bodies of water such as the Straight of Juan de Fuca, making it difficult to form a theoretical prediction of tidal strains \citep{Roeloffs2010}. Another complication is that noise in BSM data is not well understood. The noise consists, in part, of a long-term decay resulting from the instrument equilibrating with the surrounding rock \citep{Gladwin1987}. Typically, this noise is dealt with in an ad-hoc manner by fitting and removing exponentials and low-order polynomials. We envision that the GNSS derived strain rates from this paper can be used as a reference strain for calibrating BSM data and quantify its noise.

There is potential for a more thorough analysis of the spatio-temporal noise in GNSS data, $\eta$, than what was performed in Section \ref{sec:NoiseModel}. We did not explore the spatial covariance of $\eta$, which would describe common mode noise. We are able to ignore common mode error in this study; however, for other geophysical studies based on GNSS data, such as fault slip inversions, it may be necessary to incorporate a spatially covarying noise model \citep[e.g.,][]{Miyazaki2003}. We can also improve upon the seasonal model used in this study, which consists of four spatially uncorrelated sinusoids for each station. We did not explore the spatial covariance of seasonal deformation or the temporal roughness (i.e., the number of sinusoids needed to describe the observations). The periodic Gaussian process \citep{Mackay1998} is an alternative model for seasonal deformation and is well suited for exploring the roughness of seasonal deformation.  The periodic Gaussian process has zero mean and the covariance function
\begin{equation}\label{eq:Periodic}
T(t,t') = \phi^2 \exp\left(\frac{-\sin(\pi|t - t'|)^2}{2\tau^2}\right).
\end{equation}
Realizations have annual periodicity and the roughness is controlled by $\tau$. Decreasing $\tau$ has the same effect as including higher frequency sinusoids in the seasonal model. The optimal value for $\tau$ can be found with the REML method as described in Section \ref{sec:NoiseModel}.
